Supporting Information : The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2018-10-08 Paper.Rmd
modify intro : still continue

Identifying and quantifying dependencies is the core of modelling in many applications, frequently found in the
finance and risk management. For many years, measuring and modeling dependencies has focused on the
Pearson correlation because of its ease in use. But correlation can provide little information about the underlying dependence structure in the case of asymmetric dependence because it is a measure of linear association between two random variables. Therefore, it is better to use rank-based correlation measures such as Kendall’s tau and Spearman’s rho instead of Pearson correlation. However, using copula-based approach as a dependence measure reflects more knowledge about the dependence structure instead of summarizing in just a single number. The flexibility of copulas allows us to model more complex dependence structures by isolating its marginal from its joint distribution structures, with the capability in modeling a rich variety of tail behavior, such as tail dependence versus tail independence, and reflection symmetry versus reflection asymmetry. Tail dependence occurs when the correlation between two variables increases as we get "further" in the tail (either or both) of the distribution. For example, in the upper panel of Fig. \ref{fig_cop_pedag}, different dependence structures, with same Kendall correlation $\tau = 0.5$, were shown which became dominantly lower tail dependent (lower tail - upper tail dependence being positive) from dominantly upper tail dependent (lower tail - upper tail dependence being negative) as we moved along the row. Lower panel represents the similar scenario but with stronger correlation (Kendall's $\tau = 0.8$). To implement this, we need to select candidate bivariate copulas from many existing parametric copula families. A very flexible bivariate copula can itself become a welcome candidate in the pool of bivariate copulas to be selected from. Here, throughout our studies for different datasets we used a universal pool of 16 bivariate copula families (having exclusively lower, exclusively upper, both or neither tail dependence), mentioned in the \nameref{Methods}.

Though a few examples have been found which shows copula approach to model dependencies in bioinformatics [@Kim2008], medical research [@Emura2016], a short period study of intersite dependence of rainfall data [@Serinaldi2008], a recent study on characterizing and assessing risks in drought management [@She2018] or monitoring changes in precipitation-extremes [@Goswami2018], the multivariate copula approach still, we believe, remains unexplored in ecology or related field and thus have potential scope to better capture the evolution of ecological time series data. So, in ecological perspective, when we find a dominantly lower (upper) tail dependence structure, it means both incidence are rare (abundant) synchronously. If any one of them be environmental variable and the other is an observable (like abundance of a species) then it can also explain the causal relationship between them. Here, we analyzed the dependence structure for several bivariate and multivariate ecological, community and biogeochemical datasets. In addition to a copula based \nameref{Model_selection}, we also carried out a \nameref{NPA} with three different statistics to verify our model selection based results.

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=16 cm,height=8 cm]{./Results/copula_pedagog_fig_BB1.pdf}
\caption{Pedagogical figure showing BB1 copula with varying tail-dep difference but of same Kendall's tau along each row and with same tail-dep difference but of different Kendall's tau along each column.\label{fig_cop_pedag}}
\end{center}
\end{figure}

>>>>

Identifying and quantifying dependencies is the core of modelling in many applications, frequently found in finance and risk management. For many years, measuring and modeling dependencies has focused on the
Pearson correlation because of its ease in use. But correlation can provide little information about the underlying dependence structure in the case of asymmetric dependence because it is a measure of linear association between two variables. It may be better to use rank-based correlation measures such as Kendall’s tau and Spearman’s rho instead of Pearson correlation, when the marginal distribution of the variables are highly non-Gaussian. Copula, a ranked based dependence measurement between two variables with initially some non-uniformly distributed marginals, transforms the marginals into uniform distribution when joint distribution of ranked data is plotted. For example, in the first panel of Fig. \ref{fig_cop_pedag}, we showed the joint distribution of raw data (left) and copula data (right), where the marginals of raw data variables (x and y) were gamma distributed but in copula plot those marginals had uniform distribution. Copula plot was obtained by scaling individual ranked variables by a factor $(n+1)$ where $n$ was the number of sampling points, so that the copula was confined within a unit box. However, using a "copula-based approach" as a dependence measure reflects more knowledge about the dependence structure instead of summarizing in just a single number, the linear correlation coefficient. It tells us about the tail-dependence of the variables, i.e., the association between extreme values of two variables. If smaller values of each variable are more correlated to each other, then the corresponding copula structure is said to have "lower" tail-dependence as shown in the first panel of Fig. \ref{fig_cop_pedag} copula plot (right). Similarly, an "upper" tail-dependence copula indicates stronger correlation between higher extreme variables in the raw data. The flexibility of copulas allows us to model more complex dependence structures by isolating marginal from joint distribution structures, with the capability of modeling a rich variety of tail behavior, such as tail dependence versus tail independence, and reflection symmetry versus reflection asymmetry. Tail dependence occurs when the correlation between two variables increases as we get "further" in the tail (either or both) of the distribution. For example, in the upper panel of Fig. \ref{fig_cop_pedag}, different dependence structures, with same Kendall correlation $\tau = 0.5$, are shown which vary between dominantly lower tail dependent (lower tail - upper tail dependence being positive) from dominantly upper tail dependent (lower tail - upper tail dependence being negative) as we move along the row. Lower panel represents the similar scenario but with stronger correlation (Kendall's $\tau = 0.8$). To implement this, we need to select candidate bivariate copulas from many existing parametric copula families. <!--A very flexible bivariate copula can itself become a welcome candidate in the pool of bivariate copulas to be selected from.--> Here, throughout our studies for different datasets we used a universal pool of 16 bivariate copula families (having exclusively lower, exclusively upper, both or neither tail dependence), mentioned in the \nameref{Methods}.

Though a few examples have been found which shows copula approach to model dependencies in bioinformatics [@Kim2008], medical research [@Emura2016], a short period study of intersite dependence of rainfall data [@Serinaldi2008], a recent study on characterizing and assessing risks in drought management [@She2018] or monitoring changes in precipitation-extremes [@Goswami2018], the multivariate copula approach still, we believe, remains unexplored in ecology or related field and thus have potential scope to better capture the evolution of ecological time series data. In ecological perspective, when we find a dominantly lower (upper) tail dependence structure between two species abundance timeseries, it means both incidence are rare (abundant) synchronously. If one variable is an environmental variable and another is an observable (like abundance of a species) then it can also explain the causal relationship between them. Here, we analyzed the dependence structure for several bivariate and multivariate ecological, community and biogeochemical datasets. In addition to a copula based \nameref{Model_selection}, we also carried out a \nameref{NPA} with three different statistics to verify our model selection based results.

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=4 cm,height=4 cm]{./Results/rawdata.pdf}
\includegraphics[width=4 cm,height=4 cm]{./Results/copdata.pdf}\\
\includegraphics[width=16 cm,height=8 cm]{./Results/copula_pedagog_fig_BB1.pdf}
\caption{Pedagogical figure showing BB1 copula with varying tail-dep difference but of same Kendall's tau along each row and with same tail-dep difference but of different Kendall's tau along each column.\label{fig_cop_pedag}}
\end{center}
\end{figure}

### 2018-10-08 SupportingInformation.Rmd
modify intro : still continue

@@

Next, the [**BMR data**](https://github.com/sghosh89/BIVAN/blob/master/Data/BMR/bmr-contrasts.csv) contains the trait contrasts from phylogenetically independent contrasts [@SukumaranH2010] of data reported by [@GenoudIM2017] on the basal metabolic rate (BMR) and mass of mammals species. Specifically, the "SELECT" BMR data in their database were used for those species for which data were available. A phylogeny in NEXUS [@NEXUS] format was extracted from the supplementary information of [@GenoudIM2017]. Version 4.4.0 of the DendroPy library [@SukumaranH2010] was used to calculate phylogenetic independent contrasts using Python 2.7. For each contrast, the path length spanned by the contrast (sum of the branch lengths) and the contrast for the mass, natural logarithm of the BMR and natural logarithm of the mass were written as a row in the [bmr-contrasts.csv](https://github.com/sghosh89/BIVAN/blob/master/Data/BMR/bmr-contrasts.csv) file.

@@

Absolute values of traits for each contrast are considered to construct corresponding copula for second and third datasets.

Fourth, we used **Ceder-creek data for e120** which resulted from a biodiversity experiment, carried out by Tilman and his co-warkers [@Tilman2001; @Tilman2006] over many years to collect [plant above ground biomass](https://www.cedarcreek.umn.edu/research/data/dataset?ple120) and [plant species percent cover](https://www.cedarcreek.umn.edu/research/data/dataset?pce120) data. These datasets were used to calculate plant-aboveground biomass and Shannon's diversity index (H), averaged over each plot (9m by 9m),
for year 2000.

>>>>

```{r cop_fig_intro,echo=F,results="hide",cache=T, cache.extra=list(seed,mtime("copsurrognd.R"))}
library(copula)
library(VineCopula)
library(mvtnorm)
set.seed(seed)

source("copsurrognd.R")

#x <- read.table("./Data/dataset/x.txt")
#y <- read.table("./Data/dataset/y.txt")
#mat<-cbind(x,y)
#colnames(mat)<-c("x","y")

#***constants
N<-200 #length of time series
n<-2 #number of sampling locations
numdat<-1 #the number of datasets to get from the normal copula
rho<-0.8 #normal-copula parameter for dependence between locations
#this is a covariance, see below call to rmvnorm

#***generate the starting data, which has normal copula structure
#because it comes originally from a multivariate normal, before 
#transformation
sig<-matrix(rho,n,n)
diag(sig)<-1
m<-rmvnorm(N*numdat,mean=rep(0,n),sigma=sig)
m<-array(m,c(N,numdat,n))
m<-aperm(m,c(1,3,2)) # 2nd dim is normally distributed e.g. hist(m[,1,])
gshape<-7.5
gscale<-1 #gamma parameters

cop_to_marg_trans<-function(x){
  return(qgamma(x,shape=gshape,scale=gscale)) #this is where we choose the marginals,
  #should be something skewed
}
# pnorm(m) makes normal to uniform
# then qgamma(...) makes uniform to gamma distributed
m<-cop_to_marg_trans(pnorm(m))

ccop<-claytonCopula(2,2)
csurrog<-array(NA,dim(m))
csurrog[,,1]<-copsurrognd(m[,,1],targetcop=ccop,numsurrog=1)
mat<-csurrog[,,1]

pdf("./Results/rawdata.pdf",height = 4, width = 4)
scale<-2
xhist <- hist(mat[,1], breaks=30, plot=FALSE)
yhist <- hist(mat[,2], breaks=30, plot=FALSE) 
top <- max(c(xhist$counts, yhist$counts)) 
xrange <- range(mat[,1])
yrange <- range(mat[,2])
nf <- layout(matrix(c(2,0,1,3),2,2,byrow=TRUE), c(4,1), c(1,4), TRUE) 
op<-par(mar=c(3,3,1,1),mgp=c(2, 1, 0))  
plot(mat[,1], mat[,2], xlim=xrange, ylim=yrange, xlab="x", ylab="y", cex.lab=1.5) 
op<-par(mar=c(0,3,1,1)) 
barplot(xhist$counts/scale, axes=FALSE, ylim=c(0, top), space=0) 
op<-par(mar=c(3,0,1,1)) 
barplot(yhist$counts/scale, axes=FALSE, xlim=c(0, top), space=0, horiz=TRUE)
par(op)
dev.off()

pdf("./Results/copdata.pdf",height = 4, width = 4)
copmat<-pobs(mat)
xhist <- hist(copmat[,1], breaks=30, plot=FALSE)
yhist <- hist(copmat[,2], breaks=30, plot=FALSE) 
top <- max(c(xhist$counts, yhist$counts)) 
xrange <- c(0,1)
yrange <- c(0,1)
nf <- layout(matrix(c(2,0,1,3),2,2,byrow=TRUE), c(4,1), c(1,4), TRUE) 
op<-par(mar=c(3,3,1,1),mgp=c(2, 1, 0)) 
plot(copmat[,1], copmat[,2], xlim=xrange, ylim=yrange, xlab="u=rank(x)/(n+1)", ylab="v=rank(y)/(n+1)",cex.lab=1.5) 
op<-par(mar=c(0,3,1,1)) 
barplot(xhist$counts/scale, axes=FALSE, ylim=c(0, top), space=0) 
op<-par(mar=c(3,0,1,1)) 
barplot(yhist$counts/scale, axes=FALSE, xlim=c(0, top), space=0, horiz=TRUE)
par(op)
dev.off()
```

@@

<!--
Next, the [**BMR data**](https://github.com/sghosh89/BIVAN/blob/master/Data/BMR/bmr-contrasts.csv) contains the trait contrasts from phylogenetically independent contrasts [@SukumaranH2010] of data reported by [@GenoudIM2017] on the basal metabolic rate (BMR) and mass of mammals species. Specifically, the "SELECT" BMR data in their database were used for those species for which data were available. A phylogeny in NEXUS [@NEXUS] format was extracted from the supplementary information of [@GenoudIM2017]. Version 4.4.0 of the DendroPy library [@SukumaranH2010] was used to calculate phylogenetic independent contrasts using Python 2.7. For each contrast, the path length spanned by the contrast (sum of the branch lengths) and the contrast for the mass, natural logarithm of the BMR and natural logarithm of the mass were written as a row in the [bmr-contrasts.csv](https://github.com/sghosh89/BIVAN/blob/master/Data/BMR/bmr-contrasts.csv) file.

@@

Absolute values of traits for each contrast are considered to construct corresponding copula for second and third datasets.
--> 
Fourth, we used **Ceder-creek data for e120** which resulted from a biodiversity experiment, carried out by Tilman and his co-warkers [@Tilman2001; @Tilman2006] over many years to collect [plant above ground biomass](https://www.cedarcreek.umn.edu/research/data/dataset?ple120) and [plant species percent cover](https://www.cedarcreek.umn.edu/research/data/dataset?pce120) data. These datasets were used to calculate plant-aboveground biomass and Shannon's diversity index (H), averaged over each plot (9m by 9m),
for year 2000.
